Early Warning Signals and the Prosecutor's Fallacy

### 2012-08-30 prosecutor/fallacy.tex
Cleanup
Alan's edits added

%\documentclass[authoryear,5p]{elsarticle}
\documentclass[authoryear,review,12pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
\usepackage{float}
\usepackage{subfigure}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}
\newcommand{\cb}[1]{{\it \color{darkgreen} (#1)}}

@@

\journal{\tiny } 
\begin{document}
\begin{frontmatter}
  \title{Early Warning Signals and the Prosecutor's Fallacy}
  \author[cpb]{Carl Boettiger\corref{cor1}}
  \ead{cboettig@ucdavis.edu}
  \author[esp]{Alan Hastings}
  %\author[info]{}
  %\author[davis]{}
  \cortext[cor1]{Corresponding author.}
  \address[cpb]{Center for Population Biology, 1 Shields Avenue, University of California, Davis, CA, 95616 United States.}
  \address[esp]{Department of Environmental Science and Policy, University of California, Davis}

@@

The prosecution has only observed that the probability of seeing the
evidence ($E$) they produced given a random innocent individual ($I$),
$P(E|I)$ is very small.  From this one cannot conclude that the individual
is indeed guilty, that is, that the probability the individual is innocent
given the evidence $P(I|E)$ is also very small. In a city with millions
of people, there might be several individuals who match the description
of the evidence\footnote{Mathematically $P(E|I)$ need not equal $P(I|E)$,
instead, these expressions are related by Bayes theorem,

\begin{equation}
  P(E|I) = P(I|E) \frac{P(E)}{P(I)}
\end{equation}

}. The California Supreme Court reversed the decision, and
the case became a widely recognized example of the Prosecutor's
Fallacy~\citep{Thompson1987}.  Here we explore how a similar misconception
can arise from the use of historical data to evaluate methods for
detecting early warning signals of critical transitions.

@@

\section{Methods and Results}
To investigate if early warning signals are vulnerable to this fallacy,
we simulating a system that is not driven towards a bifurcation such as
in Fig~\ref{fig:fallacy}.  This simulation approach allows us to determine whether
examining historical events is a valid way to test the utility of these
indicators.  We simulated 20,000 replicates of a stochastic individual-based
birth-death process with an Allee threshold~\citep{Courchamp2008}, which
arises from positive fitness effects at low densities.  Above the Allee
threshold the population returns to a positive equilibrium size, whereas
below the threshold the population decreases to zero. The
simulation uses the Gillespie algorithm to provide an exact implementation
of the master equation of the birth-death process,

\begin{align}
  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master}, \\
    b_n &= \frac{e K n^2}{n^2 + h^2}, \\
    d_n &= e n + a,
\end{align}

a model with a linear death rate and density-dependent birth
rate that drives the Allee effect at low densities and limits
growth at high densities.  In this model $n$ indicates the discrete
number of individuals in the population, $K$ indicates a carrying 
capacity as set by a limiting resource, $e$ a per-capita death rate, 
$a$ an additional mortality imposed on the population such as harvest,
$h$ is a parameter controlling at what population size the addition of 
more individuals switches from conferring a positive benefit on growth 
from Allee interactions $n < h$ to a negative impact on growth due to 
increased competition, $n > h$.  The key feature of this model is the
alternate stable states introduced by this effect; other functional
forms for Eq.~\eqref{master} could serve equally well for these
simulations~\citep[see \emph{e.g.}][]{Scheffer2001}.  Though this
system can be forced through a bifurcation by increasing
the death rate, in these simulations all parameters are held constant
and no bifurcation occurs.  Consequently we do not anticipate an early
warning signal of an approaching bifurcation.

@@

To test for the presence of a warning signal in these indicators we
computed values of Kendall's $\tau$ for both indicators for each of
the 266 replicates.  Kendall's $\tau$ takes values in $(-1, 1)$, and is
frequently used to identify an increasing trend ($\tau > 0 )$ in early
warning signals~\citep{Dakos2008, Dakos2011}.  The distribution
of $\tau$ values observed across these replicates is shown in
Figure~\ref{fig:indicator}.   
%The common statistical significance test used for Kendall's $\tau$
%assumes independence of the variables to achieve a normal distribution,
%this isn't satisfied for indicators coming from a time series and
%computed with a sliding window.  For a more robust comparison, 
We compare the distribution of $\tau$ from all the simulations to
the distribution conditioned on experiencing a chance transition to the
alternative stable state.  To avoid an effect of sample size the time series are all chosen
to be the same length.  
%As these examples have been selected from a system
%that is not approaching a bifurcation, an indicator that does not suffer
%from the bias of Prosecutor's fallacy should have the same distribution
%in both cases.   
Annotated, parallelised code to replicate the analysis
and produce the figures is provided in the appendix.  (Currently at
\href{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}).

@@

For each of these replicates we also take a model-based
approach, estimating parameters for an approximate linear
model of the system approaching a saddle node bifurcation, as
described~\citet{Boettiger2012b}.  In this model, the parameter $m$
describes the approach towards the saddle-node bifurcation.  Estimates
$m < 0 $ are expected in systems approaching a bifurcation, while for
stable systems $m$ should be approximately zero. None of the estimates
across the 266 simulations differed from zero in our study, hence the
model-based estimation shows no evidence of bias on data that has been
selected conditional on collapse.

@@

We have shown that systems which undergo rare sudden transitions look
statistically different from their counterparts that do not, even though
they are driven by the same stochastic process.  In particular, such
conditionally selected examples are more likely to show signs associated
with an early warning of an approaching tipping point, such as increasing
variance or increasing autocorrelation, as measured by Kendall's $\tau$.
This increases the risk of false positives -- cases in which a warning
signal being tested appears to have successfully detected an oncoming
tipping point, when in fact the example comes
instead from a stable system. Figure 2 shows the extent of this bias,
where many of the chance crashes show values of $\tau$ that are 
significantly larger than those observed in the otherwise identical
replicates that did not experience a chance transition.

@@

It seems tempting to argue that the bias towards positive detection
in historical examples is not problematic -- each of these systems did
indeed collapse, so the increased probability
of exhibiting warning signals could be taken as a successful detection.
Unfortunately this is not the case. At the moment the forecast is made,
these systems are no more likely to crash than their counterparts from
the sample that did not collapse, just as a fair coin is still a fair
coin after a string of consecutive ``heads.''  A closer look at
the patterns involved shows why common indicators
such as autocorrelation and variance can be misleading.

@@

The problem we highlight ultimately stems from the difficulty of having
only a single realization with which to examine a complex problem.
The only way to deal with this problem embodied is through replication, as
can be done in an experimental system in laboratory manipulations
such as~\citet{Drake2010, Veraart2011, Dai2012} and at the scale of whole lake
ecosystems in~\citet{Carpenter2011}.  
Experimental procedures avoid the hazard of the Prosecutor's fallacy by
generating a complete sample of replicates, rather than selecting a subset
of cases from some larger historical sample.  
%We note as well that the Prosecutors Fallacy 
%may arise in other contexts where we search historical data for
%examples containing some feature of interest.

>>>>

%\documentclass[authoryear,5p]{elsarticle}
\documentclass[authoryear,review,12pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
%\usepackage{float}
\usepackage{subfigure}
%\usepackage[pdftex]{color}
%\definecolor{darkblue}{rgb}{0,0,0.5}
%\definecolor{darkgreen}{rgb}{0,0.5,0}
%\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}
\newcommand{\cb}[1]{{\it \color{darkgreen} (#1)}}

@@

\journal{\tiny Proceedings of the Royal Society B} 
\begin{document}
\begin{frontmatter}
  \title{Early Warning Signals and the Prosecutor's Fallacy}
  \author[cpb]{Carl Boettiger\corref{cor1}}
  \ead{cboettig@ucdavis.edu}
  \author[esp]{Alan Hastings}
  %\author[info]{}
  %\author[davis]{}
  \cortext[cor1]{Corresponding author.}
  \address[cpb]{Center for Population Biology, 1 Shields Avenue, University of California, Davis, CA, 95616 United States.}
  \address[esp]{Department of Environmental Science and Policy, University of California, Davis}

@@

The prosecution has only observed that the probability of seeing the
evidence ($E$) they produced given a random innocent individual ($I$),
$P(E|I)$ is very small.  From this one cannot conclude that the individual
is indeed guilty, that is, that the probability the individual is innocent
given the evidence $P(I|E)$ is also very small. In a city with millions
of people, there might be several individuals who match the description
of the evidence.  Mathematically $P(E|I)$ need not equal $P(I|E)$,
instead, these expressions are related by Bayes theorem,

\begin{equation}
  P(E|I) = P(I|E) \frac{P(E)}{P(I)},
\end{equation}

$P(E) \ll 1$ and $P(I)\approx 1$ , so $P(E|I) \approx P(I|E) P(E)$, and consequently we cannot conclude that $P(I|E) \ll 1$ from $P(E|I) \ll 1$.  
Realizing this mistake, the California Supreme Court reversed the decision, and
the case became a widely recognized example of the Prosecutor's
Fallacy~\citep{Thompson1987}.  Here we explore how a similar misconception
can arise from the use of historical data to evaluate methods for
detecting early warning signals of critical transitions.

@@

\section{Methods and Results}
To investigate if early warning signals are vulnerable to this fallacy,
we simulate a system that is not driven towards a bifurcation such as
in Fig~\ref{fig:fallacy}.  This simulation approach allows us to determine whether
examining historical events is a valid way to test the utility of these
indicators.  We simulated 20,000 replicates of a stochastic individual-based
birth-death process with an Allee threshold~\citep{Courchamp2008}, which
arises from positive fitness effects at low densities.  Above the Allee
threshold the population returns to a positive equilibrium size, whereas
below the threshold the population decreases to zero. The model can be represented as a continuous time birth-death process where births and deaths are Poisson events whcih depend on the current density,

\begin{align}
    b(X) &= \frac{e K n^2}{n^2 + h^2}, \\
    d(X) &= e n + a,

\end{align}

%The simulation uses the Gillespie algorithm to provide an exact implementation\footnote{ of the birth-death process,

%\begin{align}
%  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master}, \\
%    b_n &= \frac{e K n^2}{n^2 + h^2}, \\
%    d_n &= e n + a,
%\end{align}

a model with a linear death rate and density-dependent birth
rate that drives the Allee effect at low densities and limits
growth at high densities.  In this model $n$ indicates the discrete
number of individuals in the population, $K$ indicates a carrying 
capacity as set by a limiting resource, $e$ a per-capita death rate (the $e$ scaling term in the birth equation allows the carrying capacity $K$ to correspond to a positive equilibrium point), 
$a$ an additional mortality imposed on the population such as harvest,
$h$ is a parameter controlling at what population size the addition of 
more individuals switches from conferring a positive benefit on growth 
from Allee interactions $n < h$ to a negative impact on growth due to 
increased competition, $n > h$.  The key feature of this model is the
alternate stable states introduced by this effect; other functional
forms for Eq.~\eqref{master} could serve equally well for these
simulations~\citep[see \emph{e.g.}][]{Scheffer2001}.  Though this
system can be forced through a bifurcation by increasing
the death rate, in these simulations all parameters are held constant
and no bifurcation occurs.  Consequently we do not anticipate an early
warning signal of an approaching bifurcation.

@@

To test for the presence of a warning signal in these indicators we
computed values of Kendall's $\tau$ for both indicators for each of
the 266 replicates.  Kendall's $\tau$ is a non-parametric measure of rank correlation frequently used to identify an increasing trend ($\tau > 0 )$ in early warning signals~\citep{Dakos2008, Dakos2011}, defined as $\tau \def \tfrac{(\text{number of concordant pairs}) - (\text{number of discordant pairs})}{\frac{1}{2} n (n-1) }$ in $n$ observations.  \footnote{A pair of observations $(x_i,y_i)$ and $(x_j, y_j)$ are concordant if $x_i > x_j$ and $y_i > y_j$ or $x_i < x_j$ and $y_i < y_j$ and discordant otherwise; equalties excepted.}  $\tau$ takes values in $(-1, 1)$.  The distribution
of $\tau$ values observed across these replicates is shown in
Figure~\ref{fig:indicator}.   
%The common statistical significance test used for Kendall's $\tau$
%assumes independence of the variables to achieve a normal distribution,
%this isn't satisfied for indicators coming from a time series and
%computed with a sliding window.  For a more robust comparison, 
We compare the distribution of $\tau$ from all the simulations to
the distribution conditioned on experiencing a chance transition to the
alternative stable state.  To avoid an effect of sample size the time series are all chosen
to be the same length.  
%As these examples have been selected from a system
%that is not approaching a bifurcation, an indicator that does not suffer
%from the bias of Prosecutor's fallacy should have the same distribution
%in both cases.   
Annotated, parallelised code to replicate the analysis
and produce the figures is provided in the appendix.

@@

For each of these replicates we also take a model-based
approach, estimating parameters for an approximate linear
model of the system approaching a saddle node bifurcation, as
described by~\citet{Boettiger2012b},

\begin{equation}
\ud X = \sqrt{ r_t } (\phi(r_t) - X_t)\ud t + \sigma\sqrt{\phi(r_t) } \ud B_t \label{LSN}
\end{equation}

In this model, the parameter $m$
describes the approach towards the saddle-node bifurcation.  Estimates
$m < 0 $ are expected in systems approaching a bifurcation, while for
stable systems $m$ should be approximately zero. None of the estimates
across the 266 simulations differed from zero in our study, hence the
model-based estimation shows no evidence of bias on data that has been
selected conditional on collapse.

@@

The attempts to detect early warning signs for critical transitions
are based on the concept of a deteriorating environment as embodied
in a changing parameter~\cite{Scheffer2009}, which is a different kind
of transition than one which is driven instead by stochasticity in an
environment which is otherwise constant and exhibiting no directional
change. When trying to use historical data to understand critical
transitions we often do not know which category, changing environment
or simply chance, an observed large change falls into.

We have shown here that systems which undergo rare sudden transitions
due to chance look statistically different from their counterparts that
do not, even though they are driven by the same stochastic process.
In particular, such conditionally selected examples are more likely to
show signs associated with an early warning of an approaching tipping
point, such as increasing variance or increasing autocorrelation,
as measured by Kendall's $\tau$.  This increases the risk of false
positives -- cases in which a warning signal being tested appears to
have successfully detected an underlying change in the system leading
to a tipping point, when in fact the example comes instead from a stable
system with no underlying change in parameters. Figure 2 shows that many
of the chance crashes show values of $\tau$ that are significantly larger
than those observed in the otherwise identical replicates that did not
experience a chance transition, thus ``detecting'' an underlying change
in the system dynamics that is not in fact present.

@@

It seems tempting to argue that this bias towards positive detection in
historical examples is not problematic -- each of these systems did
indeed collapse, so the increased probability of exhibiting warning
signals could be taken as a successful detection.  Unfortunately this
is not the case. At the moment the forecast is made, these systems are
no more likely to crash than their counterparts from the sample that did
not collapse, just as a fair coin is still a fair coin after a string of
consecutive ``heads.''  A closer look at the patterns involved shows why
common indicators such as autocorrelation and variance can be misleading.

@@

\subsection{Importance of experimental approaches}
The problem we highlight ultimately stems from the difficulty of having
only a single realization with which to examine a complex problem.
The only way to deal with this problem embodied is through replication, as
can be done in an experimental system in laboratory manipulations
such as~\citet{Drake2010, Veraart2011, Dai2012} and at the scale of whole lake
ecosystems in~\citet{Carpenter2011}.  
Experimental procedures avoid the hazard of the Prosecutor's fallacy by
generating a complete sample of replicates, rather than selecting a subset
of cases from some larger historical sample.  
%We note as well that the Prosecutors Fallacy 
%may arise in other contexts where we search historical data for
%examples containing some feature of interest.
